Supplementary Information

### 2021-06-10 doc/manuscript/ForestGEO_treerings.Rmd
small changes
word count should be right at limit (8000)

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to multiple climatic drivers and their interactions with tree size.
Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with non-linear responses prevalent (63% of relationships).
Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees showing greater sensitivity. 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.

@@

Tree DBH scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010b] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

@@

Across diverse climates and forest types (Tables `r sites`, `r species_table`), growth rates of `r n_species` tree species usually responded positively to water availability (PPT or PDF) -- at least up until the long-term mean -- 
and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Fig. `r clim_sens`).
These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: outside of the wet tropics (where there are few tree-ring records), the majority of forests are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone). 
Within warmer regions, warm winter or early spring temperatures in humid climates may advance the growing season [@keenan_net_2014] and increase annual growth [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig. `r clim_sens`).
However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
The primary mechanism underlying growth decreases at high temperatures is presumably increased evaporative demand (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
This effect occurs in addition to the effects of precipitation (Fig. `r clim_sens`), highlighting the fact that temperature and associated VPD increases limit growth even under conditions of high soil moisture [@novick_increasing_2016], and occurs over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig. `r clim_sens`).
This suggests that relatively short periods of anomalously high temperatures and evaporative demand, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.

>>>>

Tree rings provide an invaluable long-term record for understanding how climate and other drivers shape tree growth and forest productivity.
However, conventional tree-ring analysis methods were not designed to simultaneously test effects of climate, tree size, and other drivers on individual growth.
This has limited the potential to test ecologically relevant hypotheses on tree growth sensitivity to multiple climatic drivers and their interactions with tree size.
Here, we develop and apply a new method to simultaneously model non-linear effects of primary climate drivers, reconstructed tree diameter (DBH), and calendar year in generalized least squares models that account for the temporal autocorrelation inherent to each individual tree's growth.
We analyze data from `r n_trees_all` trees representing `r n_species` species at 10 globally distributed sites, showing that precipitation, temperature, DBH, and calendar year have additively, and often interactively, influenced annual growth over the past 120 years.
Growth responses were predominantly positive to precipitation (usually over $\ge$ 3-month seasonal windows) and negative to temperature (usually maximum temperature, over $\le$ 3-month seasonal windows), with non-linear responses prevalent (63% of relationships).
Climate sensitivity commonly varied with DBH (44% of cases tested), with larger trees generally showing greater sensitivity. 
Trends in ring width at small DBH were linked to the light environment under which trees established, but basal area or biomass increments consistently reached a maximum at intermediate DBH.
Accounting for climate and DBH, growth rate declined over time for 92% of species in secondary or disturbed stands, whereas growth trends were mixed in older forests.
These trends were largely attributable to stand dynamics as cohorts and stands age, which remain challenging to disentangle from global change drivers.
By providing a parsimonious approach for characterizing multiple interacting drivers of tree growth, our method reveals a more complete picture of the factors influencing growth than has previously been possible.

@@

Tree DBH scales with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010b] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

@@

Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species usually responded positively to water availability (PPT or PDF) -- at least up until the long-term mean -- 
and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Fig. `r clim_sens`).
These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: outside of the wet tropics (where there are few tree-ring records), the majority of forests are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone). 
Within warmer regions, warm winter or early spring temperatures in humid climates may advance the growing season [@keenan_net_2014] and increase annual growth [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig. `r clim_sens`).
However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
The primary mechanism underlying growth decreases at high temperatures is presumably increased evaporative demand (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
This effect occurs in addition to the effects of precipitation (Fig. `r clim_sens`), highlighting the fact that temperature and associated VPD increases limit growth even under conditions of high soil moisture [@novick_increasing_2016], and occurs over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig. `r clim_sens`).
This suggests that relatively short periods of anomalously high temperatures and evaporative demand, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.
